---
title: "HW5"
output: pdf_document
date: "2023-02-23"
---

```{r}
library(vegan)
data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
?BCI
head(BCI_xy)
head (BCI)
dim(BCI)
dim(BCI_xy)

```

#' 1) Examine if there is evidence of spatial dependence in a rare and a common species in the BCI tree dataset

```{r}
# Find out what data is rare and what is common (25% and 75% quantiles)
names(BCI)
sad <- colSums(BCI)
hist(log(sad))
quantile(log(sad), probs = c(0.25, 0.75), na.rm = FALSE)
exp(1.9459)
which(log(sad) == quantile(log(sad), 0.25))
rare_species <- 'Ficus.obtusifolia'
which(log(sad) == quantile(log(sad), 0.75))
common_species <- 'Trichilia.pallida'

BCI_species <- BCI[, c(rare_species, common_species)]

# Calculating Euclidean distances
rare_dist = dist(BCI$Tetrathylacium.johansenii, method = "euclidean")
common_dist = dist(BCI$Trichilia.pallida, method = "euclidean")
xy_dist = dist(BCI_xy, method = "euclidean")

max_dist <- max(BCI_xy) / 2

# Plotting these distances and linear models, then calculated correlations
plot(xy_dist, rare_dist)
abline(lm(rare_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, rare_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)

rare_cor = cor(xy_dist, rare_dist)
rare_cor

plot(xy_dist, common_dist)
abline(lm(common_dist~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, common_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)

common_cor = cor (xy_dist, common_dist)
common_cor

# Performing permutation tests to compare these correlations to a null distribution
rare_mantel = mantel(xy_dist, rare_dist)
rare_mantel

common_mantel = mantel(xy_dist, common_dist)
common_mantel
```
# The result of the mantel test shows r-0.08744 meaning there is a significant correlation between the spatial distance matrix (xdis = dw) and the species distance matrix (ydis = Wmat). This shows that there is evidence of spatial dependence in the rare and common species selected.

2) Build two generalized linear models to predict the abundance of the species Drypetes standleyi using the abundance of other tree species in the study site. Specifically examine the following species as predictor variables:

```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")

# create a subset of the data with the target and predictor variables
subset_df <- BCI[, c("Drypetes.standleyi", sp_ids)]

# Model 1: only include a single species as a predictor variable
model1 <- glm(Drypetes.standleyi ~ Cordia.lasiocalyx, data = subset_df, family = poisson)

# Model 2: include all of the species as predictor variables
model2 <- glm(Drypetes.standleyi ~ ., data = subset_df, family = poisson)

# view the summary of the models
summary(model1)
summary(model2)

#use function Variogram for all predictor species 
vario_prespec <- Variogram(gls_prespec, form= ~ BCI$Drypetes.standleyi + bci_sub_spat, resType = 'response')

#compare to a plot we make on our own 
res <- residuals(gls_prespec)
res_var <- dist(res)^2 * 0.5
plot(dist(BCI[, c(BCI$Drypetes.standleyi, bci_sub_spat)]), res_var)
lines(lowess(dist(BCI[, c(BCI$Drypetes.standleyi, bci_sub_spat)]), res_var), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

library(geoR)

# calculate residuals for both models
resid1 <- residuals(model1, type = "response")
resid2 <- residuals(model2, type = "response")

# calculate empirical variogram of the residuals for both models
vg1 <- Variogram(as.list(as.data.frame(coordinates(BCI))), resid1)
vg2 <- Variogram(as.list(as.data.frame(coordinates(BCI))), resid2)


# plot the variograms
plot(vg1, main = "Empirical Variogram of Residuals (Model 1)")
plot(vg2, main = "Empirical Variogram of Residuals (Model 2)")


# fit exponential model to the residuals of model 1
expfit1 <- likfit(vg1, ini.cov.pars = c(0.1, 0.1))

# fit exponential model to the residuals of model 2
expfit2 <- likfit(vg2, ini.cov.pars = c(0.1, 0.1))


# fit model 1 with spatial error term
model1_spatial <- glmmadmb(Drypetes.standleyi ~ Cordia.lasiocalyx, data = BCI, 
                            random = ~ x + y, family = "nbinom")

# fit model 2 with spatial error term
model2_spatial <- glmmadmb(Drypetes.standleyi ~ ., data = BCI, 
                            random = ~ x + y, family = "nbinom")

# compare model fit with anova
anova(model1, model1_spatial, test = "Chisq")
anova(model2, model2_spatial, test = "Chisq")

```

Did including the spatial error term have a large impact on the coefficients of the model?

Did including the spatial error terms significantly improve model fit (use function anova to carry out model comparison)?

Explain why you did or did not observe a difference in the influence of adding the spatial error term between the two models.